---
title: "FCF Clasicos - A data driven answer"
author: "Juan C. Sanchez-Arias"
date: "14/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Of "El Clasico", "Derby", and other tales

The FIFA World Cup is approaching, well sort of. The qualifying season has re-started and I have been trying my earnest to stay up-to-date to what is happening with my gloriosa Seleccion Colombia, after all, it is my home country and the its soil still remembers that wonderful long distance bendy shot from high school (at least I like to think this is the case).

While watching news videos on the subject, I stumbled upon a very interesting discussion from the folks at [F90 at ESPN Fans](https://www.youtube.com/watch?v=bEC7NtQQXi8): Is the Colombia *vs.* Ecuador a "Clasico"? After a heated debate, they anchors couldn't come up with a clear answer, not even the "data guy" from the show.

Dataset source: ["International football results from 1872 to 2021" (An up-to-date dataset of over 40,000 international football results)](https://www.kaggle.com/martj42/international-football-results-from-1872-to-2017/discussion/208198).

```{r set-up, include = FALSE}
# Setting up workspace and libraries
setwd("~/GitHub/futbolito")
library(tidyverse)
library(cowplot)
library(ggalt)

futbol <- read_csv("results.csv")
```

## Wrangling data

```{r clasicos, echo=TRUE}

# fcf_home <- futbol %>% filter(home_team == "Colombia") %>% group_by(away_team) %>% count(sort = TRUE)
# 
# fcf_away <- futbol %>% filter(away_team == "Colombia") %>% group_by(home_team) %>% count(sort = TRUE)

# Coding FCF away/home status per game
fcf_total <- futbol %>% mutate(
  fcf_status = case_when
  (home_team == 'Colombia' ~ 'home', 
    away_team == 'Colombia' ~ 'away')) %>% 
  na.omit()

# Adding  rival column to ease downstream operations
fcf_total <- fcf_total %>% mutate(
  rival = case_when(
    home_team == 'Colombia' ~ away_team,
    away_team == 'Colombia' ~ home_team
  )
)

# Let's find FCF's Clasicos!
fcf_total %>% group_by(rival) %>% count(sort = TRUE) %>% head(9) %>% knitr::kable()

sa <- c("Argentina", "Bolivia", "Brazil", "Chile", "Ecuador", "Paraguay", "Peru", "Uruguay", "Venezuela")

fcf_total_sa <- fcf_total %>% filter(rival %in% sa)
```

So according to this dataset that includes all international games played from `r format((min(futbol$date)), format = "%Y")` to `r format((max(futbol$date)), format = "%Y")`, Colombia's Men National Team Top 5 "Clasicos" games are against Peru, Ecuador, Paraguay, and Uruguay.

## Let's plot FCF historical records

```{r plots, include=TRUE, echo=FALSE}
# 
# fcf_total_sa %>% group_by(rival, fcf_status) %>% count(sort = TRUE) %>%
#   ggplot(aes(x = reorder(rival, desc(n)), y = n, fill = fcf_status)) + geom_col() +
#   ylab("Game count") + xlab(element_blank()) +
#   scale_fill_brewer(palette = 'Blues') + theme_bw() + theme(legend.title = element_blank())
# 

fcf_total_sa <- fcf_total_sa %>% mutate(
  goaldiff_fcf = case_when(
    home_team == 'Colombia' ~ home_score - away_score,
    away_team == 'Colombia' ~ away_score - home_score),
  win_lost = case_when(
    goaldiff_fcf > 0 ~ "Win",
    goaldiff_fcf < 0 ~ "Draw",
    goaldiff_fcf == 0 ~ "Loss"
  )
)

fcf_total_sa$win_lost <- factor(fcf_total_sa$win_lost, levels = c("Win", "Draw", "Loss"))

# Other (better) way to plot
fcf_total_sa %>% filter(tournament != "Bolivarian Games",
                        tournament !="Brazil Independence Cup",
                        tournament != "Gold Cup") %>% ggplot(aes(x = reorder(rival, -table(rival)[rival]), 
                            fill = win_lost)) + geom_bar(alpha = 0.4) +
  ylab("Total game count") + xlab(element_blank()) +
  scale_fill_manual(values = c("yellow", "blue", "red")) + theme_bw() +
  labs(title = "FCF games between 1938 - 2021",
       caption = "Last update: 2021-09-09\ntwitter @juan_sanar") +
  theme(legend.title = element_blank(), text = element_text(family = "sans")) + facet_wrap(~tournament)
```


```{r plots, include=TRUE, echo=FALSE}

fcf_total_sa %>% filter(tournament != "Bolivarian Games",
                        tournament !="Brazil Independence Cup",
                        tournament != "Gold Cup") %>% ggplot(aes(x = reorder(rival, -table(rival)[rival]), 
                            fill = win_lost)) + geom_bar(alpha = 0.4) +
  ylab("Total game count") + xlab(element_blank()) +
  scale_fill_manual(values = c("yellow", "blue", "red")) + theme_bw() +
  labs(title = "FCF games between 1938 - 2021",
       caption = "Competitions: Copa America, FIFA World Cup Qualification, FIFA World Cup, Friendly Games\nLast update: 2021-09-09") +
  theme(legend.title = element_blank(), text = element_text(family = "sans"))

```
```{r}
p1 <- fcf_total_sa %>% filter(tournament != "Bolivarian Games",
                        tournament !="Brazil Independence Cup",
                        tournament != "Gold Cup") %>% ggplot(aes(x = reorder(rival, -table(rival)[rival]), 
                            fill = win_lost)) + geom_bar(alpha = 0.4) +
  ylab("Total game count") + xlab(element_blank()) +
  scale_fill_manual(values = c("yellow", "blue", "red")) + theme_bw() +
  labs(title = "FCF games between 1938 - 2021",
       caption = "Competitions: Copa America, FIFA World Cup Qualification, FIFA World Cup, Friendly Games\nLast update: 2021-09-09") +
  theme(legend.position = "none", text = element_text(family = "sans"))

fcf_total_sa$win_lost <- factor(fcf_total_sa$win_lost, levels = c("Win", "Draw", "Loss"))

# Other (better) way to plot
p2 <- fcf_total_sa %>% filter(tournament != "Bolivarian Games",
                        tournament !="Brazil Independence Cup",
                        tournament != "Gold Cup") %>% ggplot(aes(x = reorder(rival, -table(rival)[rival]), 
                            fill = win_lost)) + geom_bar(alpha = 0.4) +
  ylab("Total game count") + xlab(element_blank()) +
  scale_fill_manual(values = c("yellow", "blue", "red")) + theme_bw() +
  labs(title = "FCF games between 1938 - 2021 per competition",
       caption = "Last update: 2021-09-09\ntwitter: @juan_sanar") +
  theme(legend.title = element_blank(), 
        text = element_text(family = "sans"),
        axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~tournament)

plot_grid(p1, p2)

```

```{r goals, include=TRUE, echo=FALSE}

p3 <- fcf_total_sa %>% filter(tournament != "Bolivarian Games",
                        tournament !="Brazil Independence Cup",
                        tournament != "Gold Cup")  %>% 
  ggplot(
    aes(reorder(rival, -table(rival)[rival]), goaldiff_fcf, 
        colour = as.numeric(format(date, format = '%Y')),
        size = home_score + away_score)) +
  geom_jitter(alpha = 0.5, width = 0.2) +
  geom_abline(slope = 0, intercept = 0) +
  theme(legend.position = "bottom", ) + 
  scale_color_viridis_c("Year") + 
  ylim(c(-10, 10)) +
  labs(title = "Historical FCF Goal Differential",
       size = "Total goals",
       caption = "Last update: 2021-09-09\ntwitter: @juan_sanar") +
  xlab(element_blank()) +
  ylab("Goal Differential") +
  scale_size(breaks = c(0, 2, 4, 6, 8, 10)) +
  theme_bw() +
  facet_wrap(~ tournament)
```

